\mutagen: Reliable Coverage-Guided, Property-Based Testing using Exhaustive Well-Typed Mutations

### 2021-03-30 papers/oopsla-21/main.tex
WIP

We evaluated the performance of \mutagen using two main case studies.
%
In both of them, we show how the heuristics described in Section
\ref{sec:heuristics} affect the testing performance.

The first one is a simple abstract stack machine that enforces
\emph{noninterference} using runtime checks.
%
The implementation of this case study was originally proven correct by \cite{}
in Coq, and subsequently degraded by systematically introducing 20 bugs on its
enforcing mechanism.
%
\citeauthor{lampropoulos2019coverage} used this same case study to compare
\fuzzchick against random testing approaches using na\"ive and manually-written
smart random generators.
%
In this work we replicate their results and compare them against our tool.

The second case study aims to evaluate \mutagen in a more realistic scenario,
and focuses on testing \textit{haskell-wasm}, an existing WebAssembly engine
written in Haskell of industrial strength.
%
We manually injected 10 bugs in the validator as well as 5 bugs in the
interpreter of this engine, and used the reference implementation to find them
using differential testing.
%
During this process, we quickly discovered 3 bugs and 2 discrepancies on this
engine with respect to the reference implementation.
%
All of them were reported and confirmed by the authors of \textit{haskell-wasm}.

@@

The abstract stack machine used in this case study consists of four main
elements: a program counter, a stack, and data- and instruction memories.
%
Moreover, every runtime value is labeled with its security level, i.e., L (for
``low'' or public) or H (for ``high'' or secret).
%
Labels form a trivial 2-lattice where information can either stay at the same
level, or public information can flow to secret one but not the opposite.
%
Security labels are propagated throughout the execution of the program every
time the machine executes an instruction.
%
There are eight different instructions defined as:
%
\begin{verbatim}
data Instr = Nop | Push Int | Call Int | Ret | Add | Load | Store | Halt
\end{verbatim}

\noindent Control flow is achieved using the \texttt{Call} and \texttt{Ret}
instructions, that let the program jump back and forth within the instruction
memory using specially labelled values in the stack respresenting memory
addresses.
%
The argument in the \texttt{Push} instruction represents a value to be
inserted in the stack, whereas the argument of the \texttt{Call} instruction
encodes the number of elements of the stack to be treated as arguments.
%
Then, programs are simply modeled as sequences of instructions.
%
To preserve space, we encourage the reader to refer to the work of
\citeauthor{catalin} as well as the original \fuzzchick paper for more details
about the implementation and semantics of this case study.

\subsubsection{Noninterference}

\subsection{Single-step Noninterference}

@@

WebAssembly is a popular assembly-like language designed to be an open standard
for executing low-level code in the web, although it has become increasingly
popular in standalone, non-web contexts as well.
%
WebAssembly programs are first validated and later executed in a sandboxed
environment (a virtual machine), making this language an attractive target for
virtualization in \emph{functions-as-a-service} platforms.
%
The language is relatively simple, in esence
%
\begin{inparaenum}
  \item it contains only four numerical types, representing both integers and
    IEEE754 floating point numbers of either 32 or 64 bits;
  \item values of these types are manipulated by writing functions using
    sequences of stack instructions;
  \item functions are organized in modules and must be explicitly imported and
    exported;
  \item memory blocks can be imported, exported and grown dinamically;
\end{inparaenum}
%
among others.
%
To give the reader a taste, Fig. \ref{fig:wasm:factorial} the WebAssembly text
representation of module that exports a recursive implementation of the
factorial function.

\begin{wrapfigure}{r}{0.35\textwidth}
\vspace{-10pt}
\begin{verbatim}
(module
(func $f (param f64) (result f64)
  get_local 0
  f64.const 1
  f64.lt
  if (result f64)
    f64.const 1
  else
    get_local 0
    get_local 0
    f64.const 1
    f64.sub
    call $fact
    f64.mul
  end)
(export "fact" (func $f)))
\end{verbatim}
\caption{\label{fig:wasm:factorial}Factorial in WebAssembly}
\vspace{-10pt}
\end{wrapfigure}

@@

In this work, we are interested on using \mutagen to test the two most complex
subsystems of \textit{haskell-wasm}: the \emph{validator} and the
\emph{interpreter} --- both being previously tested using a unit test suite.
%
Our tool is an attractive match for testing \textit{haskell-wasm}, as the space
of WebAssembly programs contains mostly invalid ones, and automatically derived
random generators cannot satisfy all the invariants required to produce
interesting test cases.
%
Here, we avoided spending countless hours writing an extensive property-based
specification to mimic the reference WebAssembly specification.
%
Instead, we take advantage of the readily available reference implementation via
differential testing.
%
In this light, our testing properties assert that any result produced by
\textit{haskell-wasm} matches that of the reference implementation.

@@

We want to remark that, although here for simplicity we only focus on finding
false positives, in a realistic test suite, one would also want to test for
false negatives, i.e., when a module is valid and \textit{haskell-wasm} rejects
it.
%
This can be easily done by inverting the direction of the implication
(\texttt{<==}) in the property above.
%
However, the resulting property is much slower, as every tested module will be
serialized and run against the reference implementation.

@@

Testing the WebAssembly interpreter is substantially more complicated than
testing the validator, as it entails running actual programs.
%
To achieve this, we need the generated test cases to comply a with stable
interface that can be invoked both by \textit{haskell-wasm} and the reference
WebAssembly implementation.

@@

This property instantiates the module stub using the input function and its type
signature, and uses it to invoke both the \textit{haskell-wasm} and reference
implementation interepreters with the provided arguments.
%
Then, the property asserts whether their results are equivalent.
%
Interestingly, equivalence in this context does not imply equality,
non-deterministic operations in WebAssembly like NaN reinterpretations can
produce different equivalent results (as exposed by the discrepancy \#5 in Table
\ref{table:wasm:bugs}), and our equivalence relation needs to take that into
acount.
%
Notice that we additionally set a 20ms timeout to discard potentially diverging
programs with infinite loops.

@@

@@

\subsection{Grammar-based Fuzzing}

@@

\subsection{Random Data Generation}

@@

\subsection{Exhaustive Bounded Testing}

>>>>

We evaluated the performance of \mutagen using two main case studies.
%
The first one is a simple abstract stack machine that enforces
\emph{noninterference} using runtime checks.
%
The implementation of this case study was originally proven correct by \cite{}
in Coq, and subsequently degraded by systematically introducing 20 bugs on its
enforcing mechanism.
%
\citeauthor{lampropoulos2019coverage} used this same case study to compare
\fuzzchick against random testing approaches using na\"ive and manually-written
smart random generators.
%
In this work, we replicate their results and compare them against our tool.
%
Worth mentioning, since this case study was originally implemented in Coq, we
first translated it to Haskell in order to run \mutagen on its test suite.

The second case study aims to evaluate \mutagen in a more realistic scenario,
and focuses on testing \textit{haskell-wasm}, an existing WebAssembly engine
written in Haskell of industrial strength.
%
We manually injected 10 bugs in the validator as well as 5 bugs in the
interpreter of this engine, and used the reference implementation to find them
via differential testing.
%
During this process, we quickly discovered 3 bugs and 2 discrepancies on this
engine with respect to the reference implementation.
%
All of them were reported and later confirmed by the authors of
\textit{haskell-wasm}.

@@

The abstract stack machine used in this case study consists of four main
elements: a program counter, a stack, and data- and instruction memories.
%
Moreover, every runtime value is labeled with a security level, i.e., L (for
``low'' or public) or H (for ``high'' or secret).
%
Labels form a trivial 2-lattice where information can either stay at the same
level, or public information can flow to secret one but not the opposite
(represented using the familiar $\sqsubseteq$ binary operator).
%
Security labels are propagated throughout the execution of the program every
time the machine executes an instruction.
%
There are eight different instructions defined as:
%
\begin{verbatim}
data Instr = Nop | Push Int | Call Int | Ret | Add | Load | Store | Halt
\end{verbatim}

\noindent Control flow is achieved using the \texttt{Call} and \texttt{Ret}
instructions, that let the program jump back and forth within the instruction
memory using specially labelled values in the stack respresenting memory
addresses.
%
The argument in the \texttt{Push} instruction represents a value to be
inserted in the stack, whereas the argument of the \texttt{Call} instruction
encodes the number of elements of the stack to be treated as arguments.
%
Then, programs are simply modeled as sequences of instructions.
%
Finally, machine states can be modeled using a 4-tuple \textit{(pc, stk, m, i)}
that represents a particular configuration of the program counter, stack and
data- and instruction memories, respectively.
%
To preserve space, we encourage the reader to refer to the work of
\citeauthor{catalin} as well as the original \fuzzchick paper for more details
about the implementation and semantics of this case study.

\subsubsection{Single-Step Noninterference (SSNI)}

This abstract machine is designed to enforce noninterference, which is based on
the notion of \emph{indistinguishability}.
%
Intuitively, two machine states are indistinguishable from each other if they
only differ on secret data.
%
Using this notion, the particular variant of noninterference we are interested
in this work is called \emph{single-step noninterference}.
%
In simple terms, this property asserts that, given two indistinguishable
machines states, running a single instruction on both machines brings them to
resulting states that are again indistinguishable.

The tricky part about this property is, of course, satisfying its sparse
precondition: we need to generate two valid indistinguishable machine states in
order to even proceed to execute the next instruction.
%
As demonstrated by \citeauthor{lampropoulos2019coverage}, generating two
independent machine states using \quickcheck has virtually no chances of
producing valid indistinguishable ones.
%
However, having the mutation mechanism available, we can use a clever trick: we
can obtain a pair of valid indistinguishable machine states by generating a
single valid machine state (something still hard but much easier than before),
and then producing a similar mutated copy.
%
By doing this, we have a much higher chance of producing two almost identical
states that pass the sparse precondition.

\subsubsection{IFC Rules}

In this abstract machine, the enforced IFC rules are implemented using a rule
table indexed by the operation that the machine is about to perform.
%
Such table contains: the dynamic check that the abstract machine needs to
perform in order to enforce the IFC policy, along with the corresponding
security label of the program counter and the operation result after the
operation is executed.
%
For instance, to execute the \texttt{Store} operation (which stores a value in a
memory pointer) the machine needs to check that both the label of the program
counter and the label of the pointer together ``can flow'' to the label of the
destination memory cell.
%
If this condition is not met, the machine then halts as indication of a
violation in the IFC policy.
%
After this check, this operation overwrites the value at the destination cell
and updates its label with the maximum sensibility of the involved labels.
%
In the rule table, this looks as follows:

\begin{center}
\footnotesize
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Operation}
& \textbf{Precondition Check}
& \textbf{Final PC Label}
& \textbf{Final Result Label} \\
\hline
\texttt{Store}
& $l_{pc} \vee l_{p} \sqsubseteq l_{v}$
& $l_{pc}$
& $l_{v'} \vee l_{pc} \vee l_{p}$ \\
\hline
\end{tabular}
\end{center}

Where $l_{pc}$, $l_{p}$, $l_{v}$, and $l_{v'}$ represent the labels of: the
program counter, the memory pointer, and the old a new values stored in that
memory cell.
%
The symbol $\vee$ simply denotes the join of two labels, i.e., the
\emph{maximum} of their sensibilities.

In this case study, we systematically introduce several bugs on the IFC
enforcing mechanism by removing or weaking the checks stored in this rule table.

@@

WebAssembly is a popular assembly-like language designed to be an open standard
for executing low-level code in the web, although it has become increasingly
popular in standalone, non-web contexts as well.
%
WebAssembly programs are first validated and later executed in a sandboxed
environment (a virtual machine), making this language an attractive target for
virtualization in \emph{functions-as-a-service} platforms.
%
The language is relatively simple, in esence
%
\begin{inparaenum}
  \item it contains only four numerical types, representing both integers and
    IEEE754 floating point numbers of either 32 or 64 bits;
  \item values of these types are manipulated by writing functions using
    sequences of stack instructions;
  \item functions are organized in modules and must be explicitly imported and
    exported;
  \item memory blocks can be imported, exported and grown dinamically;
\end{inparaenum}
%
among others.
%
%% To give the reader a taste, Fig. \ref{fig:wasm:factorial} the WebAssembly text
%% representation of module that exports a recursive implementation of the
%% factorial function.

@@

%% \begin{wrapfigure}{r}{0.35\textwidth}
%% \vspace{-10pt}
%% \begin{verbatim}
%% (module
%% (func $f (param f64) (result f64)
%%   get_local 0
%%   f64.const 1
%%   f64.lt
%%   if (result f64)
%%     f64.const 1
%%   else
%%     get_local 0
%%     get_local 0
%%     f64.const 1
%%     f64.sub
%%     call $fact
%%     f64.mul
%%   end)
%% (export "fact" (func $f)))
%% \end{verbatim}
%% \caption{\label{fig:wasm:factorial}Factorial in WebAssembly}
%% \vspace{-10pt}
%% \end{wrapfigure}

In this work, we are interested on using \mutagen to test the two most complex
subsystems of \textit{haskell-wasm}: the \emph{validator} and the
\emph{interpreter} --- both being previously tested using a unit test suite.
%
Our tool is an attractive match for testing \textit{haskell-wasm}, as the space
of WebAssembly programs that can be respresented using its AST contains mostly
invalid ones, and automatically derived random generators cannot satisfy all the
invariants required to produce interesting test cases.
%
Here, we avoided spending countless hours writing an extensive property-based
specification to mimic the reference WebAssembly specification.
%
Instead, we take advantage of the readily available reference implementation via
differential testing.
%
In this light, our testing properties assert that any result produced by
\textit{haskell-wasm} matches that of the reference implementation.

@@

We want to remark that, although here for simplicity we only focus on finding
false positives, in a realistic test suite, one would also want to test for
false negatives, i.e., when a module is valid and \textit{haskell-wasm} rejects
it.
%
This can be easily done by inverting the direction of the implication
(\texttt{<==}) in the property above.
%
However, the resulting property is much slower, as every tested module will
always be serialized and run against the reference implementation.

@@

Testing the WebAssembly interpreter is substantially more complicated than
testing the validator, since it requires running actual programs.
%
To achieve this, we need the generated test cases to comply a with stable
interface that can be invoked both by \textit{haskell-wasm} and the reference
WebAssembly implementation.

@@

This property instantiates the module stub using the input function and its type
signature, and uses it to invoke both the \textit{haskell-wasm} and reference
implementation interepreters with the provided arguments.
%
Then, the property asserts whether their results are equivalent.
%
Interestingly, equivalence in this context does not imply equality.
%
Non-deterministic operations in WebAssembly like NaN reinterpretations can
produce different equivalent results (as exposed by the discrepancy \#5 in Table
\ref{table:wasm:bugs}), and our equivalence relation needs to take that into
acount.
%
Notice that we additionally set a 20ms timeout to discard potentially diverging
programs with infinite loops.

@@

In both case studies, we additionally show how the heuristics described in
Section \ref{sec:heuristics} affect the testing performance by disabling them
when using our tool.

@@

\paragraph{Grammar-based Fuzzing}

@@

\paragraph{Random Data Generation}

@@

\paragraph{Exhaustive Bounded Testing}
